---
title: "STAT 184 Activity #10 : UC Berkeley Sex-Based Admission Discrimination"
format:
  html:
    code-link: true
    code-fold: true
    code-tools: true
editor: visual
author: Stephanie the Fourth of Halifax (assigned pseudonym)
---

## Prompt

The year is 1973. UC Berkeley has been sued for sex-based admissions discrimination. Specifically, the plaintiff claims that women are discriminated against in admissions policies. The court has provided admissions data to you and tasked you with answering the question — if a student applies to the school, will their gender play a significant role in their chance of admission?

(We are keeping the analysis limited to this dataset)

## Goal

To find out if gender plays a significant role in a student's chance of admission into UC Berkley in 1973.

## Data Inventory

First, we will load the necessary dataset and libraries to perform data analysis and visualization.

```{r}
#| message: false
#load the data, other useful packages
url <- "https://waf.cs.illinois.edu/discovery/berkeley.csv"
data <- read.csv(url)

library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
```

## Exploratory Data Analysis (EDA)

We analyze the dataset using the head() function.

```{r}
#EDA
head(data)
summary(data)
```

We see that each row in the dataset is an applicant, with details about the year they applied, major they applied for, their gender and their admission decision. We can also see that there are 12763 total applicants for the year 1973.

## Admission Rates by Gender

We will further calculate the admission rates (calculated as a proportion) based on gender and compare the two in tabular and graphical form.

```{r}
#| label: fig-Admission-Rates-by-Gender-barplot
#| fig-cap: "Admission Rates by Gender"
# Calculate the admission rates by gender
table(data$Gender, data$Admission)

admission_rates_gender <- data %>%
  group_by(Gender) %>%
  summarize(
    Total = n(),
    Accepted = sum(Admission == "Accepted"),
    Admission_Rate = Accepted / Total
  )

# Plotting the admission rates by gender
ggplot(admission_rates_gender, aes(x = Gender, y = Admission_Rate)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Admission Rates by Gender", 
       x = "Gender", 
       y = "Admission Rate") +
  theme_minimal()

```

We see a clear disparity in the proportion of male vs female applicants admitted. Male applicants approximately have a 0.45 chance of being accepted, whereas female applicants approximately have a 0.35 chance in being accepted.

NOTE: We can only make this claim for a ceteris paribus assumption, where all applicants are considered to be of 'equal caliber', and differ only by gender.

## Admission Rates by Major

Next, we will generate a plot that can help visualize the admission rates by major.

```{r}
#| label: fig-Admission-Rates-by-Major-barplot
#| fig-cap: "Admission Rates by Major"
#Admissions rate by major
admission_rate_major <- data %>%
  select(Major, Admission) %>%
  group_by(Major) %>%
  summarize(
    Total = n(),
    Accepted = sum(Admission == "Accepted"),
    Admission_Rate = Accepted / Total
  )
ggplot(admission_rate_major, aes(x = Major, y = Admission_Rate)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Admission Rates by Major", 
       x = "Major", 
       y = "Admission Rate") +
  theme_minimal()
```

## Applications Received Grouped by Gender and Major

The following graph outlines the gender distribution of the total applications received based on the majors.

```{r}
#| message: false
#| label: fig-Applications-by-Gender-and-Major-barplot
#| fig-cap: "Total Applications by Major and Gender"
#Applications by Gender, Major
applications_gender_major <- data %>%
  select(Major, Gender) %>%
  group_by(Major, Gender) %>%
  summarize(
    Total = n()
  )
ggplot(applications_gender_major, aes(x = Major, y = Total, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Applications by Major and Gender", 
       x = "Major", 
       y = "Applications") +
  theme_minimal()
```

We see that there are a higher number of male applicants for most majors, with the exception of majors C and E.

## Admission Rates Grouped by Gender and Major

We further calculate the admission rates broken down by gender and major. This will help us tell if this trend of higher proportion of admissions for male applicants is common across all majors.

```{r}
#| message: false
#| label: fig-Admission-Rates-by-Gender-Major-barplot
#| fig-cap: "Admission Rates by Major and Gender"
#See admission rates broken down by gender and major
admission_rates_gender_major <- data %>%
  group_by(Major, Gender) %>%
  summarize(
    Total = n(),
    Accepted = sum(Admission == "Accepted"),
    Admission_Rate = Accepted / Total
  )

head(admission_rates_gender_major)

ggplot(admission_rates_gender_major, aes(x = Major, y = Admission_Rate, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Admission Rates by Major and Gender", 
       x = "Major", 
       y = "Admission Rate") +
  theme_minimal()

```

We see that based on majors, the gender based admission preference seems to be reversed for most majors. For 4 out of the 7 majors (A, B, D and F), female applicants have higher admittance rates into these programs, whereas male applicants have a higher admittance rate for majors C, E and Other.

NOTE: We can only make this claim for a ceteris paribus assumption, where all applicants are considered to be of 'equal caliber', and differ only by gender and major they apply to.

## Gender Difference in Admission Rates by Major

Finally, we will plot a graph that can help us visualize the extent of disparity in the admission rates between male and female applicants across each major.

```{r}
#| label: fig-gender-diff-admission-rates-by-major
#| fig-cap: "Difference Between Male and Female Admission Rates by Major"
#See the admissions rate gender difference by major
admission_rate_diff <- admission_rates_gender_major %>%
  select(Major, Gender, Admission_Rate) %>%
  pivot_wider(names_from = Gender, values_from = Admission_Rate) %>%
  mutate(Difference = M - F)

ggplot(admission_rate_diff, aes(x = Major, y = Difference, fill = Difference > 0)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("dark green", "skyblue"), 
                    labels = c("Female Higher", "Male Higher")) +  # Red for below 0, skyblue for above 0
  labs(title = "Difference Between Male and Female Admission Rates by Major",
       x = "Major", 
       y = "Difference (Male - Female)") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black", linewidth = 1) +  # Add horizontal dotted line at y = 0
  theme_minimal() +
  theme(legend.title = element_blank())
```

The plot above illustrates that admission preferences vary by major and tend to favor one gender over the other. The plaintiff's claim is true for specific majors—particularly, majors 'C', 'E' and 'Other', where male applicants appear to have a higher likelihood of admission compared to female applicants. However, in comparing the overall admission rates, female applicants generally show a higher probability of being admitted than male applicants. This is evident from the combined height of the green bars (representing female applicants), which noticeably exceeds that of the blue bars (representing male applicants).

To summarize, gender does play a significant role in an applicant's chance at getting admitted into UC Berkley in the year 1973. This can be confirmed by @fig-gender-diff-admission-rates-by-major, where given a specific major, there is a significant difference in the admission rates between male and female applicants. If gender had no influence on admission rates, the height difference between bars for male and female applicants would ideally be zero.

NOTE: We can only make this claim for a ceteris paribus assumption, where all applicants are considered to be of 'equal caliber', and differ only in their gender and major they apply to.
